---
title: "Network Modelling Assignment 2"
subtitle: "Task 1: Geometrically Weighted Terms in ERGMs"
author: "Group 19 | Semeli Ergatoudi, Katharina Krenzer, Duarte Santos Albuquerque"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

```{r libraries}
# stocnet suite
library(manynet)
library(netrics)   # degree/density functions moved here in manynet 2.0.0
library(autograph)

# sna suite
library(network)
library(ergm)

# plotting and presentation
library(ggplot2)
library(patchwork)
library(texreg)
library(broom)
library(dplyr)
library(tidyr)

set.seed(42)
DATA <- "../Glasgow/"
```

# Data Preparation

```{r load-data}
# Read adjacency matrices with manynet
friendship <- read_matrix(paste0(DATA, "f1.csv"), header = FALSE)

# Read node attributes
attributes <- read.csv(paste0(DATA, "demographic.csv"), header = TRUE)
alcohol     <- read.csv(paste0(DATA, "alcohol.csv"),     header = TRUE)
attributes$alcohol <- alcohol$Wave.1

# Basic descriptives using manynet / netrics
nvertices <- net_nodes(friendship)
nvertices

nedges <- net_ties(friendship)
nedges

density <- net_by_density(friendship)   # moved to netrics in manynet 2.0.0
density

outdegree <- node_by_outdegree(friendship, normalized = FALSE)
mean(outdegree)
sd(outdegree)

indegree <- node_by_indegree(friendship, normalized = FALSE)
mean(indegree)
sd(indegree)
```

```{r degree-plot, fig.cap="Out-degree and in-degree distributions of the Glasgow Wave 1 network."}
plot(outdegree) + plot(indegree)
```

```{r network-object}
# Add a name column so join_nodes can match by node index
attributes$name <- 1:nvertices

# Create network object for ERGM estimation (sna suite)
friendship_sna <- friendship |>
  mutate_nodes(name = 1:nvertices) |>
  join_nodes(attributes) |>
  as_network()

friendship_sna
```

---

# Part (1): Change Statistic of `gwodegree`

## (i) Analytical Derivation

The GWD network statistic is defined as:

$$
s_{\text{GWD}}(x,\,\alpha) \;=\; e^{\alpha}\sum_{k=1}^{n}\!\Big\{1-(1-e^{-\alpha})^k\Big\}\,D_k^{\text{out}}(x)
$$

where $D_k^{\text{out}}(x)$ is the number of nodes with exactly out-degree $k$.

**Adding a directed edge to a sender node with current out-degree $k$:**

- $D_k^{\text{out}}$ decreases by 1 (the node leaves degree class $k$).
- $D_{k+1}^{\text{out}}$ increases by 1 (the node enters degree class $k+1$).

The change statistic is therefore:

$$
\begin{aligned}
\delta_{\text{GWD}}(k,\alpha)
  &= s_{\text{GWD}}(x+\text{edge},\,\alpha) - s_{\text{GWD}}(x,\,\alpha) \\[4pt]
  &= e^{\alpha}\Big[\big\{1-(1-e^{-\alpha})^{k+1}\big\}
     - \big\{1-(1-e^{-\alpha})^{k}\big\}\Big] \\[4pt]
  &= e^{\alpha}\Big[(1-e^{-\alpha})^{k} - (1-e^{-\alpha})^{k+1}\Big] \\[4pt]
  &= e^{\alpha}\,(1-e^{-\alpha})^{k}\,\underbrace{\big[1-(1-e^{-\alpha})\big]}_{=\,e^{-\alpha}}
   = \boxed{\;(1-e^{-\alpha})^{k}\;}
\end{aligned}
$$

## (ii) Plots of the Change Statistic

```{r change-stat-setup}
delta_GWD <- function(k, alpha) (1 - exp(-alpha))^k

alpha_vals <- c(-1, -0.5, 0, 0.1, 0.2, 0.3, 0.5, log(2), 1, 2, 3)
alpha_labs <- c("-1", "-0.5", "0", "0.1", "0.2", "0.3",
                "0.5", "ln2",  "1", "2",   "3")

# Use the range of observed out-degrees
k_max   <- max(outdegree)
k_range <- 0:k_max

df_plot <- do.call(rbind, lapply(seq_along(alpha_vals), function(i) {
  data.frame(
    k     = k_range,
    delta = delta_GWD(k_range, alpha_vals[i]),
    alpha = factor(alpha_labs[i], levels = alpha_labs),
    group = ifelse(alpha_vals[i] < 0, "negative", "non-negative")
  )
}))
```

### Negative decay values (one figure per value)

```{r fig-neg-m1, fig.cap="Change statistic for alpha = -1."}
df_sub <- df_plot |> filter(alpha == "-1")
ggplot(df_sub, aes(x = k, y = delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(colour = "steelblue", linewidth = 0.9) +
  geom_point(colour = "steelblue", size = 2.5) +
  labs(title    = expression("Change statistic," ~ alpha == -1),
       subtitle = "Negative decay value",
       x = "Current out-degree k",
       y = expression(delta[GWD](k, alpha))) +
  theme_bw(base_size = 13)
```

```{r fig-neg-m05, fig.cap="Change statistic for alpha = -0.5."}
df_sub <- df_plot |> filter(alpha == "-0.5")
ggplot(df_sub, aes(x = k, y = delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(colour = "tomato", linewidth = 0.9) +
  geom_point(colour = "tomato", size = 2.5) +
  labs(title    = expression("Change statistic," ~ alpha == -0.5),
       subtitle = "Negative decay value",
       x = "Current out-degree k",
       y = expression(delta[GWD](k, alpha))) +
  theme_bw(base_size = 13)
```

### Non-negative decay values (one combined figure)

```{r fig-nonneg, fig.cap="Change statistic for all non-negative decay values. Different colours and linetypes distinguish alpha values."}
df_nonneg <- df_plot |> filter(group == "non-negative")

ggplot(df_nonneg, aes(x = k, y = delta, colour = alpha, linetype = alpha)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_colour_brewer(palette = "Set1", name = expression(alpha)) +
  scale_linetype_discrete(name = expression(alpha)) +
  labs(title = "Change statistic - non-negative decay values",
       x = "Current out-degree k",
       y = expression(delta[GWD](k, alpha))) +
  theme_bw(base_size = 13)
```

## (iii) Main Takeaways

The change statistic $\delta_{\text{GWD}}(k,\alpha)=(1-e^{-\alpha})^k$ reveals how the decay
parameter $\alpha$ governs the marginal contribution of an additional out-edge to an
already-popular sender node.

**Non-negative $\alpha$.**
When $\alpha>0$ the base $(1-e^{-\alpha})\in(0,1)$, so the change statistic is always positive
and decreases *monotonically* with $k$.
Adding an edge from a node with $k=0$ always contributes the maximum value of 1, regardless
of $\alpha$, while adding from a high-degree node contributes progressively less.
The rate of this decay is entirely determined by $\alpha$:

- *Small* $\alpha$ (e.g.\ $\alpha=0.1$): the base is close to 0, so $\delta$ drops almost
  immediately to near-zero after $k=0$.  Only ties sent by near-isolated nodes influence the
  statistic; hubs are practically invisible.
- *Large* $\alpha$ (e.g.\ $\alpha=2$ or $\alpha=3$): the base is close to 1, so $\delta$
  remains substantial even at high $k$.  Both low- and high-degree senders contribute almost
  equally, and the statistic is sensitive to the entire out-degree spectrum.
- *$\alpha=\ln 2$*: the base equals $\frac{1}{2}$, giving a clean geometric halving with each
  additional out-edge --- a natural "neutral" reference.

When $\alpha=0$ the change statistic is 0 for all $k>0$: adding an edge from any already-active
sender has no effect on the statistic whatsoever.

**Negative $\alpha$.**
The base $(1-e^{-\alpha})<0$, so $\delta$ *alternates in sign* (positive for even $k$,
negative for odd $k$) and grows in absolute magnitude.
This has no natural interpretation in terms of a degree distribution; the statistic oscillates
rather than capturing a smooth preferential-attachment or equalisation mechanism.

**Summary.**
The decay parameter acts as a "bandwidth" on the out-degree axis.
Larger $\alpha$ extends the effective range over which additional ties are counted, making the
statistic more sensitive to hubs; smaller positive $\alpha$ concentrates the statistic on
low-degree nodes.
Negative $\alpha$ produces counter-intuitive alternating behaviour and should be avoided in
substantive applications.

## (iv) Choosing $\alpha$ for an Activity Effect

An *activity* effect means that nodes who are already highly active (high out-degree) are *more*
likely to send further nominations.
The goal is for the `gwodegree` statistic to be *driven by* high-out-degree nodes, so that its
coefficient $\theta_{\text{GWD}}$ can capture this self-reinforcing dynamic.

The change statistic $(1-e^{-\alpha})^k$ is smallest for large $k$ when $\alpha$ is small,
and approaches 1 for all $k$ as $\alpha\to\infty$.
For `gwodegree` to be sensitive to hubs we need $\delta$ to remain meaningfully large at the
high end of the observed out-degree range.

In the Glasgow Wave 1 network the maximum out-degree is `r k_max`.
At $\alpha=0.3$: $\delta(6,\,0.3)=(1-e^{-0.3})^6\approx 0.0003$ -- the sixth tie is essentially
invisible.
At $\alpha=1$: $\delta(6,\,1)\approx 0.063$; at $\alpha=2$: $\approx 0.46$; at $\alpha=3$:
$\approx 0.78$.

Therefore, to model an activity effect where *high out-degree nodes drive the statistic*,
**$\alpha$ should be chosen large** (at least $\alpha=1$, preferably $\alpha=2$ or $\alpha=3$
for this network).
With a large, positive $\theta_{\text{GWD}}$ and a large $\alpha$, the model assigns higher
probability to networks in which already-active senders continue to accumulate further
nominations --- a "rich-get-richer" / Matthew effect in out-degree.

Negative $\alpha$ must be avoided: the alternating-sign behaviour of $\delta$ makes
$\theta_{\text{GWD}}$ uninterpretable and estimation prone to degeneracy.

---

# Part (2): Effect of the `gwodegree` Coefficient on the Out-Degree Distribution

## (i) M1: Edges-Only Reference Model

```{r m1, cache=TRUE}
M1 <- ergm(friendship_sna ~ edges,
           control = control.ergm(seed = 42))
summary(M1)
```

## (ii) M2: Fix Decay = 0.3, Vary `gwodegree` Coefficient

The `gwodegree` coefficient is fixed via `offset()` while only the `edges` parameter is
re-estimated, ensuring the expected network density matches the observed density.

```{r m2a, cache=TRUE}
M2a <- ergm(
  friendship_sna ~ edges + offset(gwodegree(0.3, fixed = TRUE)),
  offset.coef = c(-2),
  control     = control.ergm(seed = 42)
)
summary(M2a)
```

```{r m2b, cache=TRUE}
M2b <- ergm(
  friendship_sna ~ edges + offset(gwodegree(0.3, fixed = TRUE)),
  offset.coef = c(2),
  control     = control.ergm(seed = 42)
)
summary(M2b)
```

**Model coefficient table:**

```{r coef-table, results='asis'}
screenreg(list(M1, M2a, M2b),
          custom.model.names = c("M1", "M2a (theta=-2)", "M2b (theta=+2)"),
          caption = "Estimated edges coefficients. gwodegree coefficients are fixed via offset.")
```

## (iii) Simulate 100 Networks and Plot Out-Degree Distributions

Following the practical, we use `ergm::simulate_formula()` with the `coef` argument to fix
parameter values.

```{r simulate, cache=TRUE}
NSIM <- 100

# simulate_formula() with coef fixes all parameter values
sim_M1  <- simulate_formula(
  friendship_sna ~ edges,
  coef    = coef(M1),
  nsim    = NSIM,
  output  = "network"
)

sim_M2a <- simulate_formula(
  friendship_sna ~ edges + offset(gwodegree(0.3, fixed = TRUE)),
  coef    = c(coef(M2a)["edges"], -2),
  nsim    = NSIM,
  output  = "network"
)

sim_M2b <- simulate_formula(
  friendship_sna ~ edges + offset(gwodegree(0.3, fixed = TRUE)),
  coef    = c(coef(M2b)["edges"], 2),
  nsim    = NSIM,
  output  = "network"
)
```

```{r outdeg-data}
# Compute node count by out-degree for each simulated network
# (simulated networks are 'network' objects -> use sna::degree)
outdeg_counts <- function(sim_list, label) {
  k_max_obs <- max(outdegree)
  do.call(rbind, lapply(seq_len(NSIM), function(i) {
    degs <- sna::degree(sim_list[[i]], cmode = "outdegree")
    tab  <- tabulate(degs + 1L, nbins = k_max_obs + 1L)
    data.frame(outdegree = 0:k_max_obs, count = tab, sim = i, model = label)
  }))
}

df_sim <- bind_rows(
  outdeg_counts(sim_M1,  "M1: edges only"),
  outdeg_counts(sim_M2a, "M2a: theta_GWD = -2"),
  outdeg_counts(sim_M2b, "M2b: theta_GWD = +2")
) |>
  mutate(model = factor(model,
                        levels = c("M1: edges only",
                                   "M2a: theta_GWD = -2",
                                   "M2b: theta_GWD = +2")))

# Observed node counts by out-degree
obs_tab <- data.frame(
  outdegree = 0:max(outdegree),
  count     = tabulate(outdegree + 1L, nbins = max(outdegree) + 1L)
)
```

```{r violin-plot, fig.height=10, fig.cap="Violin plots of node count by out-degree across 100 simulated networks. Black diamonds show observed counts."}
ggplot(df_sim, aes(x = factor(outdegree), y = count, fill = model)) +
  geom_violin(scale = "width", alpha = 0.65, colour = NA) +
  geom_point(
    data        = obs_tab,
    aes(x       = factor(outdegree), y = count),
    inherit.aes = FALSE,
    shape = 23, size = 3, fill = "black", colour = "white"
  ) +
  facet_wrap(~ model, ncol = 1) +
  scale_fill_manual(
    values = c("M1: edges only"       = "#377EB8",
               "M2a: theta_GWD = -2"  = "#E41A1C",
               "M2b: theta_GWD = +2"  = "#4DAF4A")
  ) +
  labs(title    = "Out-degree distribution: 100 simulated networks vs. observed",
       subtitle = "Black diamonds = observed node counts",
       x = "Out-degree", y = "Number of nodes") +
  theme_bw(base_size = 13) +
  theme(legend.position = "none",
        strip.text      = element_text(face = "bold"))
```

## (iv) Comparison and Interpretation

The three models share the same observed density by construction (the `edges` coefficient is
re-estimated in M2a/M2b to compensate for the `gwodegree` offset), yet they produce strikingly
different out-degree distributions.

**M1 -- edges only (reference).**
With only the `edges` term the model is a directed Bernoulli random graph in which each dyad is
independently connected with a constant probability $p=\text{logistic}(\hat\theta_{\text{edges}})$.
This produces a Binomial -- and for small $p$ approximately Poisson -- out-degree distribution.
In the Glasgow network ($n=129$, density $\approx 0.027$) the simulated distributions are
compact, centred around 3--4, with very little mass above 6.
The observed counts fall broadly within the simulated range, confirming the simple random
graph as a valid baseline.

**M2a ($\theta_{\text{GWD}}=-2$, decay $=0.3$).**
A negative `gwodegree` coefficient *penalises* the GWD statistic.
Because the change statistic $(1-e^{-0.3})^k$ is largest for $k=0$ and drops rapidly,
penalising this term discourages ties sent from low-degree nodes while being less averse to
ties from high-degree ones.
The result is a more *heterogeneous*, right-skewed out-degree distribution: fewer nodes remain
at degree 0 or 1, and more mass appears at moderate-to-high degrees.
The distribution is broader than M1 and exhibits a heavier tail.

**M2b ($\theta_{\text{GWD}}=+2$, decay $=0.3$).**
A positive `gwodegree` coefficient *rewards* the GWD statistic.
Because the reward is greatest for ties from low-degree senders, the model strongly favours
nodes sending only a few ties and discourages further accumulation.
The simulated distributions are markedly more *concentrated* than M1: most nodes cluster at a
common moderate out-degree (2--4) with almost no nodes at 0 or at the upper tail.
The variance of the out-degree distribution is substantially reduced compared to both M1 and M2a.

**Comparison with a random network of the same density.**
The Bernoulli baseline (M1) already produces a relatively narrow near-Poisson distribution.
$\theta_{\text{GWD}}<0$ broadens and right-skews this distribution, introducing *activity
heterogeneity*: some students nominate many friends while others nominate few.
$\theta_{\text{GWD}}>0$ compresses it further, enforcing *homogeneous activity* across students.
In substantive terms, a positive `gwodegree` coefficient captures a tendency toward equal
participation, while a negative one allows hubs to emerge among senders.
